Weighted Model Counting Without Parameter Variables

### 2021-05-13 doc/SAT_paper/paper.tex
Cleanup

One approach to this problem considers weighted clauses and probabilistic
semantics based on Markov networks \cite{DBLP:conf/uai/GogateD10}. However, with
a new representation comes the need to invent new encodings and inference
algorithms. Our work is similar in spirit in that it introduces a new
representation for computational problems but can reuse recent WMC algorithms
based on pseudo-Boolean function manipulation, namely, \textsf{ADDMC}
\cite{DBLP:conf/aaai/DudekPV20} and \textsf{DPMC} \cite{DBLP:conf/cp/DudekPV20}.
Furthermore, we identify sufficient conditions for transforming a WMC instance
into our new format. As many WMC inference algorithms such as
\textsf{Ace}, \textsf{c2d}
\cite{DBLP:conf/ecai/Darwiche04}, and \textsf{miniC2D}
\cite{DBLP:conf/ijcai/OztokD15} work by compilation to tractable representations
such as arithmetic circuits, deterministic, decomposable negation normal form
\cite{DBLP:journals/jancl/Darwiche01}, and sentential decision diagrams (SDDs)
\cite{DBLP:conf/ijcai/Darwiche11}, another way to avoid parameter variables
could be via direct compilation to a more convenient representation. Direct
compilation of Bayesian networks to SDDs has been investigated
\cite{DBLP:conf/ecsqaru/ChoiKD13}. However, SDDs only support weights on
literals, and so are not expressive enough to avoid the issue. To the best of
the authors' knowledge, neither approach
\cite{DBLP:conf/ecsqaru/ChoiKD13,DBLP:conf/uai/GogateD10} has a publicly
available implementation.

@@

We begin with an overview of some notation and terminology. Throughout the
paper, we use set-theoretic notation for many concepts in logic. A \emph{clause}
is a set of literals that are part of an implicit disjunction. Similarly, a
\emph{formula} in CNF is a set of clauses that are part of an implicit
conjunction. We identify a \emph{model} with a set of variables that correspond
to the positive literals in the model (and all other variables are the negative
literals of the model). We can then define the \emph{cardinality} of a model as
the cardinality of this set. For example, let $\phi = (\neg a \lor b) \land a$
be a propositional formula over variables $a$ and $b$. Then an equivalent
set-theoretic representation of $\phi$ is $\{ \{ \neg a, b \}, \{ a \} \}$. Any
subset of $\{ a, b \}$ is an interpretation of $\phi$, e.g., $\{ a, b \}$ is a
model of $\phi$ (written $\{ a, b \} \models \phi$) of cardinality two, while
$\emptyset$ is an interpretation but not a model. We can now formally define
WMC.

@@

It is also worth noting that there was no observable difference in the width of
the project-join tree used by DPMC \cite{DBLP:conf/cp/DudekPV20} (which is
equivalent to the treewidth of the primal/Gaifman graph of the input formula)
before and after applying \cref{alg:transformation}---the observed performance
improvement is more likely related to the variable ordering heuristic used by
ADDs.\footnote{The data on this (along with the implementation of
  \cref{alg:transformation}) is available at
  \url{https://github.com/dilkas/wmc-without-parameters}.}

>>>>

One approach to this problem considers weighted clauses and probabilistic
semantics based on Markov networks \cite{DBLP:conf/uai/GogateD10}. However, with
a new representation comes the need to invent new encodings and inference
algorithms. Our work is similar in spirit in that it introduces a new
representation for computational problems but can reuse recent WMC algorithms
based on pseudo-Boolean function manipulation, namely, \textsf{ADDMC}
\cite{DBLP:conf/aaai/DudekPV20} and \textsf{DPMC} \cite{DBLP:conf/cp/DudekPV20}.
Furthermore, we identify sufficient conditions for transforming a WMC instance
into our new format. As many WMC inference algorithms
\cite{DBLP:conf/ecai/Darwiche04,DBLP:conf/ijcai/OztokD15} work by compilation to
tractable representations such as arithmetic circuits, deterministic,
decomposable negation normal form \cite{DBLP:journals/jancl/Darwiche01}, and
sentential decision diagrams (SDDs) \cite{DBLP:conf/ijcai/Darwiche11}, another
way to avoid parameter variables could be via direct compilation to a more
convenient representation. Direct compilation of Bayesian networks to SDDs has
been investigated \cite{DBLP:conf/ecsqaru/ChoiKD13}. However, SDDs only support
weights on literals, and so are not expressive enough to avoid the issue. To the
best of the authors' knowledge, neither approach
\cite{DBLP:conf/ecsqaru/ChoiKD13,DBLP:conf/uai/GogateD10} has a publicly
available implementation.

@@

We begin with an overview of some notation and terminology. We use $\land$,
$\lor$, $\neg$, $\Rightarrow$, and $\Leftrightarrow$ to denote conjunction,
disjunction, negation, material implication, and material biconditional,
respectively. Throughout the paper, we use set-theoretic notation for many
concepts in logic. A \emph{clause} is a set of literals that are part of an
implicit disjunction. Similarly, a \emph{formula} in CNF is a set of clauses
that are part of an implicit conjunction. We identify a \emph{model} with a set
of variables that correspond to the positive literals in the model (and all
other variables are the negative literals of the model). We can then define the
\emph{cardinality} of a model as the cardinality of this set. For example, let
$\phi = (\neg a \lor b) \land a$ be a propositional formula over variables $a$
and $b$. Then an equivalent set-theoretic representation of $\phi$ is $\{ \{
\neg a, b \}, \{ a \} \}$. Any subset of $\{ a, b \}$ is an interpretation of
$\phi$, e.g., $\{ a, b \}$ is a model of $\phi$ (written $\{ a, b \} \models
\phi$) of cardinality two, while $\emptyset$ is an interpretation but not a
model. We can now formally define WMC.

@@

It is also worth noting that there was no observable difference in the width of
the project-join tree used by DPMC (which is equivalent to the treewidth of the
primal/Gaifman graph of the input formula \cite{DBLP:conf/cp/DudekPV20}) before
and after applying \cref{alg:transformation}---the observed performance
improvement is more likely related to the variable ordering heuristic used by
ADDs.\footnote{The data on this (along with the implementation of
  \cref{alg:transformation}) is available at
  \url{https://github.com/dilkas/wmc-without-parameters}.}
